---
title: "Preparing data for SPI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing data for SPI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(cli.width = 80, width = 80)
library(spi)
```

This article explains what data SPI needs and how to prepare them for the direct SPI (`spi_direct()`) or the model-based SPI (`spi_index()`).

## 1. What you need

The direct SPI uses annual NPAFP counts and annual population under 15. The model-based SPI uses monthly NPAFP counts, annual population under 15 and district boundaries.

The input requirements are:

| | Direct SPI, `spi_direct()` | Model-based SPI, `spi_index()` |
|---|---|---|
| NPAFP cases | One row per district and year | One row per district and month, including a zero count for months with no cases |
| Population under 15 | One row per district and year | One row per district and year |
| Region (for example province) | Optional | Not used |
| District boundaries | Not needed | Required, as an `sf` polygon layer with matching district identifiers |
| INLA | Not needed | Required |

## 2. Four things to get right

### Use the same district identifier in every table

Use a stable code, such as the POLIS district GUID (`adm2_guid`), rather than the district name. Spelling differences and name changes can prevent records from matching when tables are joined by name.

<details>
<summary>Check that the identifiers match</summary>

```{r ids}
ids_cases <- unique(synth_surveillance$cases$adm2_guid)
ids_pop <- unique(synth_surveillance$population$adm2_guid)

# districts in the case data with no population
setdiff(ids_cases, ids_pop)
```

The result is empty here, so every district identifier in the case data also appears in the population table.

</details>

### Keep districts and years with zero cases

**A missing row and zero cases are not necessarily the same thing.** If a district was under surveillance but reported no NPAFP cases in a year, retain that district-year with a zero count.

Supply a separate population table covering every district-year to be analysed. `spi_direct()` treats a district-year with a population record but no case row as zero reported cases. If you use one combined table, include a row with the population and a zero case count for each such district-year.

### Check the population

Population is used to calculate expected cases, so an abrupt change can affect the SPI. `spi_direct()` checks each district's population history. The following code lists districts flagged for review:

```{r population-qc, eval = FALSE}
spi_results$population_qc |>
  dplyr::filter(population_qc != "ok")
```

A flag is a reason to check the population estimate, not to replace it automatically. A large change may reflect migration, displacement, a revised estimate or a district boundary change.

### Review changes in district geography

A new district may have no earlier records of its own. For a split or merged district, earlier records may cover a different area. Stable identifiers help with name changes, but they do not resolve changes in geography. Where possible, use a table linking old and new districts (a crosswalk) or prepare all years using the same district boundaries.

## 3. Data for the direct SPI

For the direct SPI, one table with one row per district and year is enough:

| district | province | year | npafp_cases | population_u15 |
|---|---|---|---|---|
| A | North | 2022 | 8 | 100 000 |
| A | North | 2023 | 10 | 105 000 |

```{r standard, eval = FALSE}
spi_results <- spi_direct(afp_data)
```

The `province` column is optional. Without it, reporting from the rest of the country provides the reference for districts with limited history.

If the population is in a separate table, pass it as `population`:

```{r separate-pop, eval = FALSE}
spi_results <- spi_direct(
  afp_counts,
  population = population
)
```

If your columns have different names, specify them in the function call:

```{r mapping, eval = FALSE}
spi_results <- spi_direct(
  afp_data,
  id_col = "adm2_guid",
  year_col = "yr",
  count_col = "npafp",
  population_col = "u15_pop",
  region_col = "province_name"
)
```

<details>
<summary>Starting from monthly counts</summary>

Sum the monthly counts for each district and year before running `spi_direct()`. This example uses the package's synthetic data and joins province names stored in the district boundary data.

```{r monthly}
provinces_harad <- sf::st_drop_geometry(synth_surveillance$boundaries) |>
  dplyr::distinct(adm2_guid, adm1_name)

afp_harad <- synth_surveillance$cases |>
  dplyr::mutate(year = lubridate::year(month)) |>
  dplyr::summarise(count = sum(count), .by = c(adm2_guid, year)) |>
  dplyr::left_join(provinces_harad, by = "adm2_guid")

res_harad <- spi_direct(
  afp_harad,
  population = synth_surveillance$population,
  id_col = "adm2_guid",
  count_col = "count",
  population_col = "pop_u15",
  region_col = "adm1_name",
  verbose = FALSE
)

res_harad
```

</details>

Use `spi_check_inputs()` with `method = "direct"` to check the table before calculating. It reports missing columns, duplicate district-years, invalid counts or populations, district-years with no row, and districts whose population series needs review:

```{r check-direct, eval = FALSE}
spi_check_inputs(afp_data, method = "direct")
```

[The direct SPI](spi-direct.html) shows how the calculation works and how to inspect the result for one district.

## 4. Data for the model-based SPI

The model-based SPI needs monthly NPAFP counts, annual population under 15 and district boundaries. Use the same district identifier in all three inputs.

| district_id | month | count |
|---|---|---|
| A | 2022-01-01 | 1 |
| A | 2022-02-01 | 0 |

Include one row for every district-month, with a zero count for months with no cases. Start with all districts in the boundary file so that districts with no case records are included. Create a row for each district and month, then join the case counts:

```{r grid, eval = FALSE}
# every district in the boundary file
districts <- tibble::tibble(
  district_id = unique(boundaries$district_id)
)

# every month in the period to analyse
months <- tibble::tibble(
  month = seq(as.Date("2022-01-01"), as.Date("2026-12-01"), by = "month")
)

# one row for every district and month
district_months <- dplyr::cross_join(districts, months)

# add the counts; a district-month with no count becomes zero
cases <- district_months |>
  dplyr::left_join(monthly_counts, by = c("district_id", "month")) |>
  dplyr::mutate(count = dplyr::coalesce(count, 0L))
```

Use `spi_check_inputs()` to check the case table, population table and district boundaries together. This check does not require INLA:

```{r check}
spi_check_inputs(
  cases = synth_surveillance$cases,
  population = synth_surveillance$population,
  shapefile = synth_surveillance$boundaries,
  id_col = "adm2_guid"
)
```

[Getting started with SPI](spi-getting-started.html) shows how to fit the model and calculate the SPI from these inputs.

## Using POLIS data?

[Using SPI with POLIS data](spi-polis.html) gives a complete workflow from the AFP line list and population data to both the direct and model-based SPI.
