---
title: "The direct SPI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The direct SPI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(cli.width = 80, width = 80)
library(spi)

num <- function(x, d = 2) {
  formatC(x, format = "f", digits = d, big.mark = " ")
}
pct <- function(x) paste0(round(100 * x), "%")

# the French example runs only when the translation service can be reached
can_translate <- tryCatch(
  requireNamespace("sntutils", quietly = TRUE) &&
    requireNamespace("gtranslate", quietly = TRUE) &&
    !identical(
      suppressMessages(
        sntutils::translate_text("Previous rate", target_language = "fr")
      ),
      "Previous rate"
    ),
  error = function(e) FALSE
)
```

```{r example-country, include = FALSE}
# An invented country of 30 districts in three provinces. We follow
# districts A, B and C; the others are simulated around a common rate.
set.seed(1)
ids <- sprintf("D%02d", 1:30)
ids[c(8, 11, 22)] <- c("A", "B", "C")
provinces <- rep(c("North", "Central", "South"), each = 10)

others <- tibble::tibble(
  district = ids,
  province = provinces,
  pop0 = round(stats::runif(30, 80000, 300000), -3),
  rate = 8e-5 * exp(stats::rnorm(30, 0, 0.3))
) |>
  dplyr::cross_join(tibble::tibble(year = 2022:2026)) |>
  dplyr::mutate(
    population_u15 = round(pop0 * 1.03^(year - 2022), -3),
    npafp_cases = stats::rpois(
      dplyr::n(), population_u15 * rate * ifelse(year == 2026, 0.9, 1)
    )
  ) |>
  dplyr::filter(!district %in% c("A", "B", "C")) |>
  dplyr::select(district, province, year, npafp_cases, population_u15)

teaching <- tibble::tribble(
  ~district, ~year, ~npafp_cases, ~population_u15,
  "A", 2022, 8, 100000, "A", 2023, 10, 105000,
  "A", 2024, 9, 110000, "A", 2025, 11, 115000,
  "A", 2026, 6, 120000,
  "B", 2025, 1, 40000, "B", 2026, 1, 42000,
  "C", 2022, 0, 250000, "C", 2023, 1, 250000,
  "C", 2024, 0, 250000, "C", 2025, 1, 250000,
  "C", 2026, 1, 255000
) |>
  dplyr::left_join(tibble::tibble(district = ids, province = provinces),
                   by = "district")

afp <- dplyr::bind_rows(others, teaching) |>
  dplyr::arrange(district, year)

# The package estimates the adjusted rates in step 5. The article works
# through the remaining arithmetic below.
res <- spi_direct(afp, first_assessment = 2026, verbose = FALSE)
row <- function(d) res$summary[res$summary$district == d, ]
a <- row("A")
b <- row("B")
c_ <- row("C")
nat <- res$national
info_a <- a$history_pop * a$reference_rate / 100000
```

The direct surveillance performance index (SPI) compares a district's reported cases with the number expected from earlier reporting, then compares that observed-to-expected ratio with the national ratio for the same year. This article works through the calculation for one district and shows how to use `spi_direct()` for all districts.

Every step, including the adjustment for districts with limited history, is arithmetic that can be checked by hand.

<style>
.spi-flow {
  display: flex;
  align-items: stretch;
  justify-content: center;
  gap: 0.25rem;
  margin: 1.5rem 0 2rem 0;
  flex-wrap: wrap;
}
.spi-flow-step {
  flex: 1 1 88px;
  max-width: 130px;
  min-width: 84px;
  padding: 0.6rem 0.35rem;
  border: 1px solid #b8b8b8;
  border-radius: 6px;
  text-align: center;
  background: #fafafa;
}
.spi-flow-step strong {
  display: block;
  margin-bottom: 0.35rem;
  font-size: 0.85rem;
}
.spi-flow-step span {
  display: block;
  font-size: 0.75rem;
  line-height: 1.3;
  color: #555;
}
.spi-flow-final {
  border: 2px solid #555;
  background: #f2f2f2;
}
.spi-flow-arrow {
  display: flex;
  align-items: center;
  font-size: 1rem;
  color: #777;
}
[data-bs-theme="dark"] .spi-flow-step {
  border-color: #555;
  background: #242422;
}
[data-bs-theme="dark"] .spi-flow-step span {
  color: #bbb;
}
[data-bs-theme="dark"] .spi-flow-final {
  border-color: #bbb;
  background: #2c2c2a;
}
@media (max-width: 700px) {
  .spi-flow-arrow {
    display: none;
  }
  .spi-flow-step {
    flex-basis: 45%;
    max-width: none;
  }
}
</style>

<p><strong>The calculation has six parts.</strong> We start with the
district's earlier reporting and finish by comparing its
observed-to-expected ratio with the national pattern.</p>

<div class="spi-flow">
<div class="spi-flow-step">
<strong>1. Earlier reporting</strong>
<span>Cases + population from earlier years</span>
</div>
<div class="spi-flow-arrow">→</div>
<div class="spi-flow-step">
<strong>2. Rate used</strong>
<span>Stabilised when history is limited</span>
</div>
<div class="spi-flow-arrow">→</div>
<div class="spi-flow-step">
<strong>3. Expected cases</strong>
<span>Current population × rate used</span>
</div>
<div class="spi-flow-arrow">→</div>
<div class="spi-flow-step">
<strong>4. Observed / expected</strong>
<span>Reported cases ÷ expected cases</span>
</div>
<div class="spi-flow-arrow">→</div>
<div class="spi-flow-step">
<strong>5. National comparison</strong>
<span>National observed ÷ expected</span>
</div>
<div class="spi-flow-arrow">→</div>
<div class="spi-flow-step spi-flow-final">
<strong>6. SPI</strong>
<span>District O/E ÷ national O/E</span>
</div>
</div>

<p>The sections below work through each part using District A.</p>

## Part I: Build the SPI by hand

### 1. What does the SPI measure?

The annual non-polio acute flaccid paralysis (NPAFP) rate is:

$$
\text{NPAFP rate} = \frac{\text{NPAFP cases}}{\text{population under 15}}
\times 100\,000
$$

In 2026, District A reported 6 NPAFP cases among 120 000 children under 15, a rate of 5.0 per 100 000. This rate tells us how many cases were reported for the size of the population.

The direct SPI uses the same cases and population, together with earlier years of data. It first asks: **given what this district reported before, how many cases would we expect now?** It compares reported cases with that expected count, then compares the district's ratio with the national ratio.

### 2. Build it step by step

The example is an invented country of 30 districts in three provinces, with one row per district and year. We follow District A and assess 2026.

```{r district-a}
afp |>
  dplyr::filter(district == "A")
```

#### Step 1. What did the district report before?

Only the years before 2026 are used: 8 + 10 + 9 + 11 = **38 cases**.

#### Step 2. Add the populations from the earlier years

Add the under-15 populations for 2022 to 2025: 100 000 + 105 000 + 110 000 + 115 000 = **430 000 child-years**. One child-year represents one child for one year. Adding annual populations accounts for both population size and the number of years covered.

#### Step 3. What was its previous reporting rate?

38 / 430 000 × 100 000 = **8.84 per 100 000**.

The same three steps in R:

```{r steps-1-3}
previous_a <- afp |>
  dplyr::filter(district == "A", year < 2026) |>
  dplyr::summarise(
    previous_cases = sum(npafp_cases),
    previous_child_years = sum(population_u15)
  ) |>
  dplyr::mutate(
    previous_rate = previous_cases / previous_child_years * 100000
  )

previous_a
```

#### Step 4. How much history supports the previous rate?

A rate based on several years and a large population tells us more about a district's earlier reporting than a rate based on one year and a small population. Few reported cases do not necessarily mean little information: a large district may have consistently reported few cases over several years. A consistently low rate can therefore be well established from the data, even though the rate itself may indicate poor absolute reporting.

#### Step 5. If information is limited, use reporting from other districts

##### Find the reference rate

In this example, each district's **reference rate** is the rate from the other districts in its province, using only years before 2026. The calculation uses this rate to help estimate expected reporting when the district's own history is limited.

$$
\text{reference rate} = \frac{\text{earlier cases in the other districts}}
{\text{earlier child-years in the other districts}} \times 100\,000
$$

```{r province-rate}
afp |>
  dplyr::filter(province == "North", district != "A", year < 2026) |>
  dplyr::summarise(
    province_cases = sum(npafp_cases),
    province_child_years = sum(population_u15)
  ) |>
  dplyr::mutate(
    province_rate = province_cases / province_child_years * 100000
  )
```

District A therefore has:

- its own previous rate: 8.84 per 100 000
- the provincial reference rate: `r num(a$reference_rate)` per 100 000

##### Put the district's history in context

We use the district's child-years from earlier years and the reference rate to measure its **historical observation**. This puts its population size and length of history into a single number, expressed as the number of cases expected at the reference rate.

$$
\text{historical observation} =
\text{earlier child-years} \times \frac{\text{reference rate}}{100\,000}
$$

This is expressed as the number of cases that would have been expected over the historical period at the reference rate.

For District A:

$$
430\,000 \times \frac{`r num(a$reference_rate)`}{100\,000}
= `r num(info_a, 1)`
$$

At the provincial reference rate, about `r round(info_a)` cases would have been expected over District A's earlier years. This is not District A's expected count for 2026. It describes how much observation lies behind its previous rate and is used when deciding how much stabilisation is needed. District A has substantial historical observation, so we would expect its own earlier reporting to contribute strongly to the rate used.

We now have the three pieces needed for the adjustment: District A's previous rate, its reference rate, and the amount of historical observation behind its previous rate.

##### Get the rate used for the expectation

To steady a rate that rests on little history, add **one case's worth of the reference rate** to the district's history: one extra case, and the child-years in which the reference rate would produce one case.

$$
\text{rate used} =
\frac{\text{earlier cases} + 1}
{\text{earlier child-years} + 100\,000 / \text{reference rate}}
\times 100\,000
$$

For District A:

$$
\frac{38 + 1}{430\,000 + 100\,000 / `r num(a$reference_rate)`} \times 100\,000
= `r num(a$stabilised_rate)`
$$

```{r rate-used}
previous_a |>
  dplyr::mutate(
    reference_rate = a$reference_rate,
    rate_used = (previous_cases + 1) /
      (previous_child_years + 100000 / reference_rate) * 100000
  )
```

This adjustment is called **stabilisation**. One added case barely changes a rate built on 38 cases, so District A's rate used stays very close to its previous rate of 8.84. For a district with little earlier observation, the added case matters more: District B's rate moves from `r num(b$history_rate)` to `r num(b$stabilised_rate)`, towards its province's `r num(b$reference_rate)`.

##### Why the amount of history matters

| District | Previous rate | Reference rate | Rate used |
|---|---|---|---|
| A | `r num(a$history_rate)` | `r num(a$reference_rate)` | `r num(a$stabilised_rate)` |
| B | `r num(b$history_rate)` | `r num(b$reference_rate)` | `r num(b$stabilised_rate)` |
| C | `r num(c_$history_rate)` | `r num(c_$reference_rate)` | `r num(c_$stabilised_rate)` |

Rates are NPAFP cases per 100 000 children under 15.

District A's rate changes little. District B has only one earlier year and 40 000 child-years, so the reference rate has more influence on its adjusted rate. District C reported only two cases, but its history covers four years and 1 million child-years. Its adjusted rate therefore remains low.

Because one case's worth is always added, a district with no cases reported in earlier years still receives a positive expected rate. A district with no earlier years takes the reference rate.

#### Step 6. Turn the rate into an expected number of cases

We now have a rate for what District A would be expected to report in 2026. To turn that rate into a number of cases, multiply it by the current population.

$$
\text{expected cases} = \text{current population} \times
\frac{\text{rate used}}{100\,000}
$$

This is the NPAFP rate calculation run in reverse. Instead of using cases and population to calculate a rate, we use the rate and population to calculate the number of cases expected.

$$
E = 120\,000 \times \frac{`r num(a$stabilised_rate)`}{100\,000}
= `r num(a$expected)`
$$

District A's expected count is therefore **`r num(a$expected)` cases**.

#### Step 7. Compare reported with expected

District A actually reported 6 cases.

$$
O/E = \frac{6}{`r num(a$expected)`} = `r num(a$oe)`
$$

Its observed-to-expected ratio is therefore **`r num(a$oe)`**. Before considering what happened nationally, District A reported about `r pct(a$oe)` of its expected count.

```{r steps-6-7}
current_a <- afp |>
  dplyr::filter(district == "A", year == 2026) |>
  dplyr::mutate(
    rate_used = a$stabilised_rate,
    expected = population_u15 * rate_used / 100000,
    observed_to_expected = npafp_cases / expected
  )

current_a |>
  dplyr::select(npafp_cases, population_u15, rate_used, expected,
                observed_to_expected)
```

#### Step 8. Compare with what happened nationally

Reporting may also have fallen nationally in 2026. To see how District A compares, we calculate expected cases for every district, then divide the total reported cases by the total expected cases. This gives the national observed-to-expected ratio.

$$
O/E_{\text{national}} = \frac{\sum O}{\sum E}
$$

```{r step-8}
national <- res$summary |>
  dplyr::summarise(
    observed = sum(observed),
    expected = sum(expected)
  ) |>
  dplyr::mutate(national_oe = observed / expected)

national
```

The SPI divides the district's ratio by the national ratio:

$$
\text{SPI} = \frac{O/E_{\text{district}}}{O/E_{\text{national}}}
= \frac{`r num(a$oe)`}{`r num(a$national_oe)`} = `r num(a$spi)`
$$

District A's SPI is therefore `r num(a$spi)`. Its observed-to-expected ratio was `r if (a$spi < 1) "lower than" else if (a$spi > 1) "higher than" else "in line with"` the national ratio.

```{r spi-a}
current_a |>
  dplyr::mutate(spi = observed_to_expected / national$national_oe) |>
  dplyr::select(district, observed_to_expected, spi)
```

### 3. The whole calculation on one page

| Question | Calculation | District A |
|---|---|---|
| How many cases were reported in earlier years? | Add the cases from earlier years | 38 |
| How much population and time does the history cover? | Add the annual populations under 15 | 430 000 child-years |
| What was the previous reporting rate? | 38 / 430 000 × 100 000 | 8.84 |
| How much observation supports the previous rate? | Earlier child-years × reference rate / 100 000 | `r round(info_a)` cases at the reference rate |
| What rate is used? | (38 + 1) / (430 000 + 100 000 / reference rate) × 100 000 | `r num(a$stabilised_rate)` per 100 000 |
| How many cases were expected? | 120 000 × `r num(a$stabilised_rate)` / 100 000 | `r num(a$expected)` |
| What is the district's observed-to-expected ratio? | 6 ÷ `r num(a$expected)` | `r num(a$oe)` |
| How did the country compare? | National observed ÷ expected | `r num(a$national_oe)` |
| What is the SPI? | `r num(a$oe)` ÷ `r num(a$national_oe)` | `r num(a$spi)` |

## Part II: Understand the result

### 4. What does the SPI mean?

| SPI | Meaning |
|---|---|
| 1.00 | The district's observed-to-expected ratio equals the national ratio |
| 0.80 | The district's observed-to-expected ratio is 20% below the national ratio |
| 0.60 | The district's observed-to-expected ratio is 40% below the national ratio |
| 1.20 | The district's observed-to-expected ratio is 20% above the national ratio |

The observed-to-expected ratio (O/E) is reported cases divided by expected cases. SPI divides the district's O/E by the national O/E for the same year. An SPI below 1 means the district's ratio is lower than the national ratio. Both ratios could still be above 1, so a low SPI does not necessarily mean the district reported fewer cases than expected.

A district can report fewer cases than expected and still have an SPI above 1, if the whole country fell further. For example, a district that reported 90% of its expected cases in a year when the country reported 75% has an SPI of 1.20:

$$
\text{SPI} = \frac{\text{district O/E}}{\text{national O/E}}
= \frac{0.90}{0.75} = 1.20
$$

The direct SPI compares current reporting with earlier reporting and with the national ratio. It does not establish that AFP cases were missed. Interpret it alongside the NPAFP rate and the other surveillance indicators. See [Interpreting SPI alongside AFP indicators](spi-interpretation.html).

### 5. Check the numbers behind the SPI

To check a district's SPI, inspect the reported cases, population, earlier reporting rate, expected cases, and national ratio used in the calculation:

| Component | District A |
|---|---|
| Current NPAFP cases | 6 |
| Current population under 15 | 120 000 |
| Current NPAFP rate | `r num(a$npafp_rate)` |
| Earlier cases | `r num(a$history_cases, 0)` |
| Earlier child-years | `r num(a$history_pop, 0)` |
| Previous rate | `r num(a$history_rate)` |
| Historical information (`history_info`) | `r a$history_info` |
| Rate used for expectation | `r num(a$stabilised_rate)` |
| Expected cases | `r num(a$expected)` |
| District observed/expected | `r num(a$oe)` |
| National observed/expected | `r num(a$national_oe)` |
| SPI | `r num(a$spi)` |

If the SPI looks surprising, check:

- Is the population plausible?
- How much information from earlier years does the district have?
- Is the expected number consistent with its earlier reporting?
- Did reporting also change nationally?

All rates in the table are per 100 000 children under 15. A district that has consistently reported few cases for its population can have an SPI near 1, so check its NPAFP rate as well.

## Part III: Let the package do the calculation

### 6. Calculate it

Everything above is calculated automatically by `spi_direct()`. With one row per district and year and these column names, one line is enough:

| district | province | year | npafp_cases | population_u15 |
|---|---|---|---|---|
| A | North | 2022 | 8 | 100 000 |
| A | North | 2023 | 10 | 105 000 |
| ... | ... | ... | ... | ... |

```{r run}
spi_results <- spi_direct(afp, verbose = FALSE)
```

Printing the result gives a short summary.

```{r print-results}
spi_results
```

By default, assessment starts three years after the first year in the data, providing three earlier years of history. This call therefore calculates results for 2025 and 2026. To calculate only the 2026 results used above, set `first_assessment = 2026`.

The `province` column is used automatically to find each district's reference rate. For a differently named region column, set `region_col` to that name. Without a region, or when the other districts in the region have no earlier cases, the reference comes from the rest of the country.

`spi_direct()` can also use a separate population table and non-standard column names; see `?spi_direct` and [Preparing data for SPI](spi-data-preparation.html).

### 7. Inspect a district

`spi_direct_explain()` shows one district's inputs and results in the same order as the worked calculation:

```{r explain}
spi_direct_explain(spi_results, "A", year = 2026)
```

Use this output to check the inputs and follow the arithmetic from the estimated rate to the final SPI. The questions in section 5 can help explain an unexpected result.

#### In another language

The labels are in English by default. Set `language` to show them in another language, for example French:

```{r explain-fr, eval = can_translate}
spi_direct_explain(spi_results, "A", year = 2026, language = "fr")
```

The labels are translated with the sntutils package, which uses Google Translate through the gtranslate package. Both packages must be installed and an internet connection is needed. The numbers and the `check_` flags are not translated.

### 8. Check the input data

The SPI uses the population to calculate expected cases, so an abrupt change in the population can move the result. `spi_direct()` checks each district's population series before calculating and returns unusual patterns in `population_qc`. It does not alter the populations you supply, and a flag does not mean the population is wrong.

```{r population}
spi_results$population_qc |>
  dplyr::filter(population_qc != "ok")
```

The flags identify different patterns:

- `check_history`: the district lacks population records for some years represented in the dataset, up to the assessment year. In this example, District B first appears in 2025.
- `check_spike`: an unusual population change is followed by an unusual change in the opposite direction within two years.
- `check_change`: another unusual population change.

A flag is a reason to check the source data. A large change can reflect real population movement, a revised population estimate, or a boundary change.

## Direct and model-based SPI

The direct SPI gives one estimate for each district and year. It shows the cases, population, earlier reporting, and adjusted rate used in the calculation, but does not provide an uncertainty interval.

The model-based SPI estimates expected reporting using patterns across districts and time and quantifies uncertainty in those expectations. This allows the precision of the resulting SPI to be assessed. See [Getting started with SPI](spi-getting-started.html) for that workflow.

## Technical details

<details>
<summary>How the stabilisation works</summary>

The rate used adds one case's worth of the reference rate to the district's history. This is the same as a weighted average of the district's previous rate and the reference rate, where the district's own rate gets the weight

$$
\frac{\text{historical observation}}{\text{historical observation} + 1}
$$

and historical observation is the number of cases expected at the reference rate over the district's earlier child-years. District A, with about `r round(info_a)` expected cases of observation, keeps `r round(100 * info_a / (info_a + 1))`% of the weight on its own rate. A district observed for one short year keeps much less. Nothing is estimated: the same rule applies to every district and every year.

The full specification is in [How SPI works](spi-methods.html#the-direct-spi). In R, open it with `vignette("spi-methods", package = "spi")`.

</details>
